Using Occupancy Theory and Hierarchical Models to Estimate Otter Population in Southeastern Minnesota \\ {\large Senior Comps Project}

### 2010-03-01 Paper.tex
Changed interval of p in CAR section
Tons of edits.

% This should be a lot prettier.
\title{Using Occupancy Theory and Hierarchical Models to Estimate Otter
Population in Southeastern Minnesota}
\date{}
\author{Chrisna Aing, Sarah Halls, Kiva Oken}
\maketitle

\section{Questions of investigation}
The North American river otter (\textit{Lontra canadensis}) is endemic to all of
Minnesota; however, populations declined precipitously in the 19th and 20th
centuries due to trapping and habitat degradation. In recent years their
populations have significantly rebounded, especially in southeastern Minnesota
where hunters would like to be able to trap otters again. In order for that to
be possible governmental organizations must be able to monitor the population
health. In addition, otter presence can indicate high quality aquatic habitat.
Due to these reasons, there is interest in monitoring changes in otter abundance
over time. Unfortunately, otters are not easy to observe. Thus, this study
investigated the potential of using wintertime aerial surveys of otter tracks on
rivers to make inferences about the abundance of otters and how it changes over
time. We used the concepts from occupancy theory to develop models to estimate
the occupancy rate in order to get a sense of otter abundance.

    \subsection{Data collection}
    During the winters of 2003 and 2004 people flew over the Whitewater, Zumbro,
    and Mississippi Rivers. When they saw what looked like an otter track, they
    pushed a button recording the GPS coordinates. They pushed the button every
    five seconds until the track ended. Flights occurred one, two, or three days
    after a snow event. In some cases, multiple observers flew over the same
    river on the same day. Surveys were conducted during multiple snow events
    throughout both winters. The rivers were then divided into 400 and 800 meter
    plots and the presence (1) or absence (0) of a waypoint in a plot on a given
    flight was determined. The Mississippi River was also divided into 1600
    meter plots.

    \subsection{Issues related to the data}
    One major issue with respect to the data is that if a track occurred near
    the boundary between two plots, a person on one flight may have recorded the
    track in one site whereas someone on a later flight during that same snow
    event may have seen the same track but recorded it in the neighboring plot.
    So while the data indicate that the observers on the two flights disagreed
    in two plots, they saw the same track. Increasing the plot size should
    decrease the magnitude of this effect.

    In addition, there is concern that the probability of an otter laying down a
    track in one site is not independent of the probability that an otter lays
    down a track in neighboring sites, violating a principle assumption of
    occupancy theory. That is, a single track can extend across multiple plots.
    This leads to dependence among the plots. It was important to quantify this
    dependence and find an optimum way to handle it.

@@

    The main reasons that people are interested in measuring occupancy are
    science or conservation and management \cite{MacKenzie2006}. This statistic
    can give insight into the way a system works. It also can be used to monitor
    animals habitats and discover trends over time in groups of animals.

    \subsection{Different ways of sampling data}
    In order to determine occupancy rate, the data is often collected in the
    form of a presence-absence survey. This means that observers visit sites and
    look for the animals themselves, tracks, vocalizations, or some other
    indicator that the animal is present. In some cases, camera traps and sound
    recording are used instead of human observers \cite{MacKenzie2006}.

    The main problem with these methods of sampling data is that it is possible
    for an animal to occupy a site, but not be detected by the observer. This
    means that the site is recorded as unoccupied, but there really is an animal
    there. This is called a false negative. Marking an animal present when there
    really is no animal is also a possible problem, but is thought to be much
    less likely and is often ignored. We will discuss this issue more in later sections.

    \subsection{Assumptions}
    As is true for all modeling, there are some assumptions made in most models of occupancy 
in order to create simple enough models \cite{MacKenzie2006}. One common assumption is no false positives, which means if a
    site is recorded as occupied we say it must actually be occupied. A second
    common assumption is that presence and detection probabilities are constant
    across sites and surveys. This simplifies the estimation of each parameter.
    A third common assumption is that there is independence of detection between
    sites. A fourth common assumption is that the surveys are being conducted on
    a closed population. This means that no animals leave the area or joins the
    area during the time of the surveys.

    There are different methods for dealing with violations of these assumptions, some 
more developed than others.  In this paper, we will only talk about attempting to get rid of the
    assumption of no false positives and the assumption of independence of
    parameters across sites.

@@

    The most simple way to measure occupancy is to simply divide the number of
    sites where an animal was found to be present by the total number of sites.
    This is called na\"ive occupancy \cite{MacKenzie2006}. It is generally
    accepted that this estimate is too low because animals are not detected at
    all sites where they are present, due to the issue of false negatives
    mentioned above. Other occupancy models attempt to do better than this
    simple estimate.

    One way to create a more accurate model is to attempt to model the biological 
proccess that is occuring.  In order to do this, we can use a Hierarchical model as a
    way to think about the parameters that affect occupancy. This allows us to
    take into account that when an animal is not detected, it could either not
    be there, or it could be there and just was not detected. This could either
    be due to a false negative, where the observer did not see the indication --
    for example they did not see the animal, track, etc. -- or it could be due
    to the fact that the animal does occupy the site but there was no indication
    of the animal at the site -- for example, no track, no vocalization, etc.

    Here is a visualization of how the Hierarchical model works:

    \begin{center}
        \includegraphics[scale=0.75]{SimpleHierarchicalModel}
    \end{center}

    From this diagram, it is apparent that we need separate parameters for the
    probability that a site is occupied, the probability that presence is
    indicated given that a site is occupied, and the probability of detection
    given that presence is indicated and the site is occupied. We will call
    these \(\psi\), \(\theta\), and \(p\), respectively.

    \subsection{Descriptive statistics}
    In 2003, three different observers flew over the rivers. Assuming that no
    tracks are laid down between individuals' flights on the same day, all three
    people should have seen the same tracks. While there is some error in
    classifying in which plot the observer saw the actual track, each observer
    should at least be seeing the same, or nearly the same, number of sites with
    tracks. However, we did not find that to be the case. If we look only at
    days that all three people observed, there appear to be major differences
    among the proportion of sites in which the observers detected a track
    (Figure \ref{obsPlots}). The difference in estimated na\"ive occupancy among 
    the
    observers indicates that the observers did not always correctly identify
    the tracks.

    \begin{figure}
        	\centering
	    \includegraphics[width=5in]{observerPlots.pdf}
	    \caption{The box plots show how the proportion of sites in which an 
	    observer saw tracks (``occupancy'') differs among the three observers in 
	    2003. Only those dates on which all three observers flew are included.}
	    \label{obsPlots}
    \end{figure}

    Correlation among the flight sequences is low. For example, average
    correlation between flights on the Mississippi in 2003 is 0.176 for 400
    meter plots. Some of this is due to the difficulties in determining to which
    site each track belongs. However, we have also found quite a bit of
    variation in the data not necessarily due to these classification issues.
    This is difficult to quantify. Correlation between flights is a bit higher 
    for
    flights on the same day or flights during the same snow event, but still not
    as high as one might hope. Correlation rarely exceeds 0.5. The proportion of 
    sites with tracks detected
    generally increases with days since snow, likely because the otters have a
    longer time to lay down a track as time after the snow event passes.

    \subsection{False positive rates}
    The lack of consistency among the flights is strong evidence that the data
    contain false positives where observers think they see a track when it does
    not exist. Not accounting for such a process can lead to major biases; as
    the number of flights increases, the occupancy rate estimate will approach
    1 \cite{Royle2006}. Unlike most occupancy models where observers assume 
    that detection of the species in question guarantees its presence \cite
    {MacKenzie2006}, we introduced a false detection process whereby an observer 
    can mark down the existence of a track when in reality there is none. This 
    false detection process can occur as long as there is no track, whether or 
    not the site is occupied.

    \subsection{Testing the independence assumption}
    \label{shat}
    Since one of the assumptions of occupancy modeling is that there is 
	independence between sites, we wanted to see if this assumption held true 
	for our data.  In order to do this, we calculated the statistic \(S\), which
	we define as the number of successive \(0\), \(1\)'s minus the number of
    successive \(1\), \(1\)'s per flight. Under the null hypothesis that each
    site is an independent Bernoulli trial with probability of a \(1 = p\) we were able
    to compute the mean and standard deviation of \(S\), which we used to
    calculate a Z-score, where \(Z = (\hat{S} - E[S])/SD[S]\). In order to compute these values, we needed to find the
    expected value and variance of each term in \(S\). These caclulations can be
	found in the appendix \ref{math}.

    Simulations of \(S\) show that the distribution of \(S\) is roughly normal.
    So using our calculations for the mean and the standard deviation, we
    computed a Z-score for each flight, where we use a plug-in estimate for
    \(p\) to be the proportion of \(1\)'s per flight. We said that \(|Z| > 2\) suggests that the assumption of
    independence between sites is violated for that flight. We found that, in
    general, alternating sites are more "independent" than all sites as measured
    by this statistic. However, alternating sites do not guarantee independence.
    For instance, six of eleven flights along the Zumbro River during 2003 still
    have significant Z-scores with alternating sites. It also appears that there 
	may be slightly more independence when larger site sizes are used, although it is
not clear if this trend is significant.  There 
	were no noticable trends for this statistic with respect to observer, days 
	since snow, or snow event.

    % Mississippi River!
    \begin{center}
    \begin{tabular}{|l|l|l|l|l|}
        \hline
        \multicolumn{5}{|l|}{\textbf{Mississippi River}} \\
        \hline
        \multirow{2}{*}{Year} & \multirow{2}{*}{Size} & \multirow{2}{*}{Flights}
        & \multicolumn{2}{|l|}{\# Sig. Z-Scores} \\
        \cline{4-5}
        & & & All Plots & Alt. Plots \\
        \hline
        \multirow{3}{*}{2003} & 402m & \multirow{3}{*}{20} & 7 & 0 \\
        \cline{2-2} \cline{4-5}
        & 804m & & 5 & 0 \\
        \cline{2-2} \cline{4-5}
        & 1608m & & 1 & 0 \\
        \hline
        \multirow{3}{*}{2004} & 402m & \multirow{3}{*}{6} & 5 & 3 \\
        \cline{2-2} \cline{4-5}
        & 804m & & 3 & 0 \\
        \cline{2-2} \cline{4-5}
        & 1608m & & 2 & 0 \\
        \hline
    \end{tabular}
    \end{center}

@@

    % Zumbro River!
    \begin{center}
    \begin{tabular}{|l|l|l|l|l|}
        \hline
        \multicolumn{5}{|l|}{\textbf{Zumbro River}} \\
        \hline
        \multirow{2}{*}{Year} & \multirow{2}{*}{Size} & \multirow{2}{*}{Flights}
        & \multicolumn{2}{|l|}{\# Sig. Z-Scores} \\
        \cline{4-5}
        & & & All Plots & Alt. Plots \\
        \hline
        \multirow{2}{*}{2003} & 402m & \multirow{2}{*}{11} & 10 & 6 \\
        \cline{2-2} \cline{4-5}
        & 804m & & 8 & 5 \\
        \hline
        \multirow{2}{*}{2004} & 402m & \multirow{2}{*}{4} & 3 & 3 \\
        \cline{2-2} \cline{4-5}
        & 804m & & 3 & 1 \\
        \hline
    \end{tabular}
    \end{center}

    We think that the lack of independence results from strings of ones where an
    otter left a track across multiple sites. When alternating sites are used,
    the strings are cut in half and the occurrence of \(1\), \(1\)'s is lower
    while the occurrence of \(0\), \(1\)'s stays about the same. It is
    noteworthy that virtually all the violations of independence are in the
    direction of more \(1\), \(1\)'s than would be expected. This reinforces our
    thoughts about successive \(1\)'s representing the same otter. We will
    discuss later whether this violation of independence actually matters in our
    model, and how this might affect the estimation of actual occupancy rates.

@@

    \subsection{Bayesian statistics}
	Most occupancy models utitlize maximum likelihood estimation, but we chose 
	to use a Bayesian approach to model the data for several reasons. 
	First, the occupancy models developed using maximum likelihood methods 
	assume that the sampled locations are a random sample from a large, 
	theoretically infinite, population.  In this study, the entire area of 
	interest was observed, so this assumption does not hold true.  Bayesian 
	models do not make such an assumption.  Second, because our model added extra
	parameters, formulating the likelihood function was cumbersome and maximizing
	it would have been difficult.  We added an additional random process, track 
	laying, and also included the possibility of false detection.  A Bayesian
	model seems much more intuitive and natural.

    Once we had a model formulated, we could then use Markov Chain Monte Carlo 
	(MCMC) to find a posterior distributions for every parameter. Due to the 
	nature of occupancy models, certain models can have multiple interpretations.
	We run into a problem of duality with the probability of correct detection
    \(p\), and the probability of false detection \(E\). These two variables
    have bimodal posterior distributions, because \(p=x,E=y,\psi=z\) is 
	equivalent in terms of likelihood to \(p=y,E=x,\psi=1-z\) due to the 
	symmetry of the model \cite{Royle2006}.  Since we know that is reasonable 
	for \(p\) to be greater than \(E\), we solved this problem by restricting 
	the distribution of \(p\) to the range \(0.6\text{ to }1\), and \(E\) to 
	\(0\text{ to }0.4\), an adaptation of the approach recommended by Royle and 
	Link \cite{Royle2006}.

@@

    WinBUGS requires a representation of a model, data, and initial values on
    which a predetermined number of iterations of the MCMC algorithm are
    executed \cite{MacKenzie2006}.
    The data we received from the
    Minnesota Department of Natural Resources was not fixed with respect to
    our covariates and required us to implement one of the three common
    strategies used to deal with irregular datasets \cite{Spiegelhalter2003}.
    Our simulations, on the other
    hand, ran on datasets with a fixed number of observers, snow events, and
    observations after each snow event, which are much easier to implement
    within WinBUGS and R.

@@

    \subsection{Our hierarchical model}
    For our model, we specified noninformative uniform \((0, 1)\) prior
    distributions for
    parameters \(\psi\) and \(\theta\), a common, natural choice given a lack of
    formal information about the prior or expert consensus \cite{MacKenzie2006}.
    We implemented models with normal and beta prior distributions on \(p\) and
    \(E\) and limited the range of \(p\) to the interval \((0.6, 1)\) and the
    range of \(E\) to the interval \((0, 0.4)\) in both models. We set the
    burn-in time for the MCMC simulation to 10,000 iterations. Each of these
    parameters are tracked and therefore are estimated. The covariates of this
    model are: observer, date, and the number of days since the last snow event.
    It was necessary to preprocess the Minnesota DNR's data to account for its
    irregularities.

    \subsection{Simulated data}
    We simulated both balanced and unbalanced datasets meant for the standard
    hierarchical model with given values of \(\psi\), \(\theta\), \(p\) for each
    observer, and \(E\) for each observer.  Balanced data sets, in this context,
    contain the same number of flights by the same observers after each snow
    event.  Naturally, unbalanced data sets do not.

    To simulate balanced data, we provided the following additional variables:
    the number of sites, the number of snow events, the number of observers, and
    the number of observations following each snow event. These variables are
    all that are required to allow us to generate data in which each snow event
    receives the same number of flights by the same number of observers, which
    greatly simplifies the accompanying hierarchical model. First, the
    simulation determines a site's occupancy based upon the given \(psi\) value.
    Then, for each day after each snow event, the simulation decides whether or
    not an otter laid down a track. For a given day after a snow event, a track
    is considered to be present if an otter lays a track down on any of the days
    after the snow event up to and including the given day.  Once that has been
    decided, the given \(p\) and \(E\) values are used to assign a 0 (not
    detected) or 1 (detected).

@@

    To determine the efficacy of the standard model over the range of possible
    \(\psi\) and \(\theta\) values, we ran simulations with several combinations
    of \(\psi\) and \(\theta\), ran the model on each simulation, and compared
    the model's estimates to the simulation's parameters. The model with beta
    prior distributions on \(p\) and \(E\) performed better than the model with
    normal prior distributions on the same parameters. A table detailing the
    model's performance with the beta prior distributions lies below:

>>>>

% This should be a lot prettier.
\title{Using Occupancy Theory and Hierarchical Models to Estimate Otter
Population in Southeastern Minnesota \\
{\large Senior Comps Project}}
\author{Chrisna Aing, Sarah Halls, Kiva Oken}
\maketitle

\section{Questions of investigation}
The North American river otter (\textit{Lontra canadensis}) is endemic to all of
Minnesota; however, populations declined precipitously in the 19th and 20th
centuries due to trapping and habitat degradation. In recent years their
populations have significantly rebounded, especially in southeastern Minnesota
where hunters would like to be able to trap otters again. In order for that to
be possible, governmental organizations must be able to monitor the health of
the population.
In addition, otter presence can indicate the quality of the aquatic habitat.
Due to these reasons, there is interest in monitoring changes in otter
population
over time. Unfortunately, otters are not easy to observe. Thus, this study
investigated the potential of using wintertime aerial surveys of otter tracks on
rivers to make inferences about the population of otters and how it changes over
time. We used concepts from occupancy theory to develop models to estimate
the occupancy rate in order to get a sense of otter population.

    \subsection{Data collection}
    During the winters of 2003 and 2004 people flew over the Whitewater, Zumbro,
    and Mississippi Rivers. When they saw what looked like an otter track, they
    pushed a button recording the GPS coordinates. They pushed the button every
    five seconds until the track ended. Flights occurred one, two, or three days
    after a snow event. In some cases, multiple observers flew over the same
    river on the same day. Surveys were conducted during multiple snow events
    throughout both winters. The rivers were then divided into 402 and 804 meter
    plots and the presence (1) or absence (0) of a waypoint in a plot on a given
    flight was determined. The Mississippi River was also divided into 1608
    meter plots.  For each river and plot size, the data set with every other
    plot removed was considered.  We refer to this as ``alternating plots''.

    \subsection{Issues related to the data}
    One major issue with respect to the data is that if a track occurred near
    the boundary between two plots, a person on one flight may have recorded the
    track in one plot whereas someone on a later flight during that same snow
    event may have seen the same track but recorded it in the neighboring plot.
    So while the data indicate that the observers on the two flights disagreed
    in two plots, they saw the same track. Increasing the plot size should
    decrease the magnitude of this effect.

    In addition, there is concern that the probability of an otter laying down a
    track in one plot is not independent of the probability that an otter lays
    down a track in neighboring plots, violating a principle assumption of
    occupancy theory. That is, a single track can extend across multiple plots.
    This leads to dependence among the plots. It was important to quantify this
    dependence and find an optimum way to handle it.

@@

    \subsection{Assumptions}
    As is true for all modeling, there are some assumptions made in most models
    of occupancy
    in order to create simple enough models \cite{MacKenzie2006}. One common
    assumption is no false positives, which means if a
    plot is recorded as occupied we say it must actually be occupied. A second
    common assumption is that presence and detection probabilities are constant
    across plots and surveys. This simplifies the estimation of each parameter.
    A third common assumption is that there is independence of detection between
    % Question: Independence of occupancy as an assumption as well?
    plots. A fourth common assumption is that the surveys are being conducted on
    a closed population. This means that no animals leave or join the
    area during the time of the surveys.

    There are different methods for dealing with violations of these
    assumptions, some more developed than others.  In this paper, we will only
    talk about attempting to eliminate the
    assumption of no false positives and the assumption of independence of
    parameters across plots.

@@

    The most simple way to measure occupancy is to divide the number of
    plots where an animal was found to be present by the total number of plots.
    This is called na\"ive occupancy \cite{MacKenzie2006}. It is generally
    accepted that this estimate is too low because observers usually do not
    detect animals at
    all plots where they are present.

    One way to create a more accurate model is to attempt to model the
    biological process that is occurring.  In order to do this, we can use a
    hierarchical model as a
    way to think about the parameters that affect occupancy. A plot can be
    designated as unoccupied for two reasons.
    There could either be
    a false negative, where the observer did not see the track,
    or
    the animal may occupy the plot but did not leave a track before a survey.

    \begin{figure}
    \begin{center}
        \includegraphics[scale=0.75]{SimpleHierarchicalModel}
        \caption{A visualization of how the hierarchical model works.}
        \label{simplehierarchicalmodel}
    \end{center}
    \end{figure}

    Figure \ref{simplehierarchicalmodel} shows that we need separate parameters
    for the
    probability that a plot is occupied, the probability that a track is laid
    down given that a plot is occupied, and the probability of detection
    given that a track is laid down and the plot is occupied. We will call
    these \(\psi\), \(\theta\), and \(p\), respectively.

\section{Exploring the data}

    \subsection{Descriptive statistics}
    In 2003, three different observers flew over the rivers. Assuming that no
    tracks are laid down between individuals' flights on the same day, all three
    people should have seen the same tracks. While there is some error in
    classifying in which plot the observer saw the actual track, each observer
    should at least be seeing the same, or nearly the same, number of plots with
    tracks. However, we did not find that to be the case. If we look only at
    days that all three people observed, there appear to be major differences
    among the proportion of plots in which the observers detected a track
    (Figure \ref{obsPlots}). The difference in estimated na\"ive occupancy among
    the
    observers indicates that the observers did not always correctly identify
    the tracks.

    \begin{figure}
        \centering
        \includegraphics[width=5in]{observerPlots.pdf}
        \caption{The box plots show how the proportion of plots in which an
        observer saw tracks (``occupancy'') differs among the three observers in
        2003. Only those dates on which all three observers flew are included.}
        \label{obsPlots}
    \end{figure}

    Correlation among the flight sequences is low. For example, average
    correlation between flights on the Mississippi in 2003 is 0.176 for 402
    meter plots. Some of this is due to the difficulties in determining to which
    plot each track belongs. However, we have also found quite a bit of
    variation in the data not necessarily due to these classification issues.
    This is difficult to quantify. Correlation between flights is a bit higher
    for
    flights on the same day or flights during the same snow event, but still not
    as high as one might expect. Correlation rarely exceeds 0.5. The proportion
    of plots with tracks detected
    generally increases with days since snow, likely because the otters have a
    longer time to lay down a track as time after the snow event passes.

    \subsection{False positive rates}
    The lack of consistency among the flights is strong evidence that the data
    contain false positives where observers think they see a track when it does
    not exist. Not accounting for such a process can lead to major biases; as
    the number of flights increases, the occupancy rate estimate will approach
    1 \cite{Royle2006}. Unlike most occupancy models where observers assume
    that detection of the species in question guarantees its presence \cite
    {MacKenzie2006}, we introduced a false detection process whereby an observer
    can mark down the existence of a track when in reality there is none. This
    false detection process can occur as long as there is no track, whether or
    not the plot is occupied.

    \subsection{Testing the independence assumption}
    \label{shat}
    Since one of the assumptions of occupancy modeling is that there is
    independence between plots, we wanted to see if this assumption held true
    for our data.  In order to do this, we calculated the statistic \(S\), which
    we define as the number of \(0\)--\(1\)'s minus the number of
    \(1\)--\(1\)'s per flight. Under the null hypothesis that each
    plot is an independent Bernoulli trial where \(p\) is the probability of a
    \(1\) we were able
    to compute the mean and standard deviation of \(S\), which we used to
    calculate a Z-score, where \(Z = (\hat{S} - E[S])/SD[S]\). In order to
    compute these values, we needed to find the
    expected value and variance of each term in \(S\). These calculations can be
    found in Appendix \ref{math}.

    Simulations of \(S\) show that the distribution of \(S\) is roughly normal.
    Using our calculations for the mean and the standard deviation, we
    computed a Z-score for each flight, where we use a plug-in estimate for
    \(p\) to be the proportion of \(1\)'s per flight. \(|Z| > 2\) suggests that
    the assumption of
    independence between plots is violated for that flight. We found that, in
    general, data with alternating plots are more ``independent'' than all plots
    as measured
    by this statistic. However, alternating plots do not guarantee independence.
    For instance, six of eleven flights along the Zumbro River during 2003 still
    have significant Z-scores with alternating plots (Table \ref{Z-scores}). It
    also appears that there
    may be slightly more independence when larger plot sizes are used, although
    it is not clear if this trend is significant. There
    were no noticeable trends for this statistic with respect to observer, days
    since snow, or snow event.

    % Mississippi River!
    \begin{table}
    \caption{Summary of \(\hat{S}\) Z-scores}
    \label{Z-scores}
    \begin{center}
    \begin{tabular}{|l|l|l|l|l|}
        \hline
        \multicolumn{5}{|l|}{\textbf{Mississippi River}} \\
        \hline
        \multirow{2}{*}{Year} & \multirow{2}{*}{Size} & \multirow{2}{*}{Flights}
        & \multicolumn{2}{|l|}{\# Sig. Z-Scores} \\
        \cline{4-5}
        & & & All Plots & Alt. Plots \\
        \hline
        \multirow{3}{*}{2003} & 402m & \multirow{3}{*}{20} & 7 & 0 \\
        \cline{2-2} \cline{4-5}
        & 804m & & 5 & 0 \\
        \cline{2-2} \cline{4-5}
        & 1608m & & 1 & 0 \\
        \hline
        \multirow{3}{*}{2004} & 402m & \multirow{3}{*}{6} & 5 & 3 \\
        \cline{2-2} \cline{4-5}
        & 804m & & 3 & 0 \\
        \cline{2-2} \cline{4-5}
        & 1608m & & 2 & 0 \\
        \hline
    \end{tabular}
    \end{center}

@@

    % Zumbro River!
    \begin{center}
    \begin{tabular}{|l|l|l|l|l|}
        \hline
        \multicolumn{5}{|l|}{\textbf{Zumbro River}} \\
        \hline
        \multirow{2}{*}{Year} & \multirow{2}{*}{Size} & \multirow{2}{*}{Flights}
        & \multicolumn{2}{|l|}{\# Sig. Z-Scores} \\
        \cline{4-5}
        & & & All Plots & Alt. Plots \\
        \hline
        \multirow{2}{*}{2003} & 402m & \multirow{2}{*}{11} & 10 & 6 \\
        \cline{2-2} \cline{4-5}
        & 804m & & 8 & 5 \\
        \hline
        \multirow{2}{*}{2004} & 402m & \multirow{2}{*}{4} & 3 & 3 \\
        \cline{2-2} \cline{4-5}
        & 804m & & 3 & 1 \\
        \hline
    \end{tabular}
    \end{center}
    \end{table}

    We think that the lack of independence results from strings of ones where an
    otter left a track across multiple plots. When alternating plots are used,
    the strings are cut in half and the occurrence of \(1\)--\(1\)'s is lower
    while the occurrence of \(0\)--\(1\)'s stays about the same. It is
    noteworthy that virtually all the violations of independence are in the
    direction of more \(1\)--\(1\)'s than would be expected. We believe that
    successive \(1\)'s could represent a single track made by the same otter. We
    will
    discuss later whether this violation of independence matters in our
    model, and how this might affect the estimation of actual occupancy rates.

@@

    \subsection{Bayesian statistics}
    Most occupancy models utilize maximum likelihood estimation, but we chose
    to use a Bayesian approach to model the data for several reasons.
    First, the occupancy models developed using maximum likelihood methods
    assume that the sampled locations are a random sample from a large,
    theoretically infinite, population.  In this study, the entire area of
    interest was observed, so this assumption does not hold true.  Bayesian
    models do not make such an assumption.  Second, because our model added
    extra
    parameters, formulating the likelihood function was cumbersome and
    maximizing
    it would have been difficult.  We added an additional random process, track
    laying, and also included the possibility of false detection.  A Bayesian
    model seems much more intuitive and natural.

    Once we had a model formulated, we could then use Markov Chain Monte Carlo
    (MCMC) to find a posterior distribution for every parameter. Due to the
    nature of occupancy models, certain models can have multiple
    interpretations.
    We run into one such duality with the probability of correct detection
    \(p\), and the probability of false detection \(E\). These two variables
    have bimodal posterior distributions, because \((p=x,E=y,\psi=z)\) is
    equivalent in terms of likelihood to \((p=y,E=x,\psi=1-z)\) due to the
    symmetry of the model \cite{Royle2006}.  Since we know that is reasonable
    for \(p\) to be greater than \(E\), we solved this problem by restricting
    the distribution of \(p\) to the range \((0.6,1)\), and \(E\) to
    \((0,0.4)\), an adaptation of the approach recommended by Royle and
    Link \cite{Royle2006}.

@@

    WinBUGS requires a representation of a model, data, and initial values on
    which a predetermined number of iterations of the MCMC algorithm are
    executed \cite{MacKenzie2006}.
    The data we received from the
    Minnesota Department of Natural Resources was not fixed with respect to
    our covariates and required us to implement one of the three common
    strategies used to deal with irregular data sets \cite{Spiegelhalter2003}.
    Our simulations, on the other
    hand, ran on data sets with a fixed number of observers, snow events, and
    observations after each snow event, which are much easier to implement
    within WinBUGS and R.

@@

    \subsection{Our hierarchical model}
    For our model, we specified noninformative uniform \((0, 1)\) prior
    distributions for
    parameters \(\psi\) and \(\theta\), a common, natural choice given a lack of
    formal information about the prior or expert consensus \cite{MacKenzie2006}.
    We implemented models with normal and beta prior distributions on \(p\) and
    \(E\) and limited the range of \(p\) to the interval \((0.6, 1)\) and the
    range of \(E\) to the interval \((0, 0.4)\) in both models. We set the
    burn-in time for the MCMC simulation to 10,000 iterations. Each of these
    parameters are tracked for 10,000 additional iterations and are estimated
    by the mean of the posterior distribution. The covariates of this
    model are the observer and the number of days since the last snow event.
    It was necessary to preprocess the Minnesota DNR's data to account for its
    irregularities.

    \subsection{Simulated data}
    We simulated both balanced and unbalanced data sets meant for the standard
    hierarchical model with given values of \(\psi\), \(\theta\), \(p\) for each
    observer, and \(E\) for each observer.  Balanced data sets, in this context,
    contain the same number of flights by the same observers after each snow
    event.  Naturally, unbalanced data sets do not.

    To simulate balanced data, we provided the following additional variables:
    the number of plots, the number of snow events, the number of observers, and
    the number of observations following each snow event. These variables are
    all that are required to allow us to generate data in which each snow event
    receives the same number of flights by the same number of observers, which
    greatly simplifies the accompanying hierarchical model. First, the
    simulation determines a plot's occupancy based upon the given \(\psi\)
    value.
    Then, for each day after each snow event, the simulation decides whether or
    not an otter laid down a track. For a given day after a snow event, a track
    is considered to be present if an otter lays a track down on any of the days
    after the snow event up to and including the given day. Once that has been
    decided, the given \(p\) and \(E\) values are used to assign a 0 (not
    detected) or 1 (detected).

@@

    To determine the efficacy of the standard model over the range of possible
    \(\psi\) and \(\theta\) values, we ran simulations with several combinations
    of \(\psi\) and \(\theta\), ran the model on each simulation, and compared
    the model's estimates to the simulation's parameters (Table \ref{Simulated
    data}). The model with beta
    prior distributions on \(p\) and \(E\) performed better than the model with
    normal prior distributions on the same parameters.  Therefore, we limit
    further discussion to the beta model.

    \begin{table}
    \caption{Beta model estimates from simulated data. All simulations
    ran with 70 plots, 5 snow events, 3 observers, and for 3 days after each
    snow event. The 3 observers had \(p\) values of 0.70, 0.80, 0.90 and \(E\)
    values of 0.05, 0.20, and 0.30, respectively. Green rows represent values of
    \(\psi\) and \(\theta\) for which the model exhibits good convergence and
    the 95\% confidence interval captures the actual \(\psi\) value.  Yellow
    rows represent values for which our model exhibits either mediocre
    convergence or a bad confidence interval.  Red rows represent values for
    which our model does not converge.}
    \label{Simulated data}
    \begin{center}
    \begin{tabular}{|l|l|l|l|}
        \hline
            \(\psi\) & \(\theta\) & Convergence & 95\% CI (\(\psi\)) \\
        \hline
        \multirow{5}{*}{0.20}
            & \cellcolor{Green}0.20 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.112, 0.305) \\
            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.142, 0.341) \\
            & \cellcolor{Green}0.60 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.098, 0.284) \\
            & \cellcolor{Green}0.80 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.097, 0.281) \\
            & \cellcolor{Yellow}1.00 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.117, 0.302) \\
        \hline
        \multirow{5}{*}{0.40}
            & \cellcolor{Yellow}0.20 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.388, 0.721) \\
            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.270, 0.494) \\
            & \cellcolor{Green}0.60 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.287, 0.512) \\
            & \cellcolor{Red}0.80 & \cellcolor{Red}Bad &
              \cellcolor{Red}(0.277, 0.507) \\
            & \cellcolor{Yellow}1.00 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.279, 0.503) \\
        \hline
        \multirow{5}{*}{0.60}
            & \cellcolor{Yellow}0.20 & \cellcolor{Yellow}Good &
              \cellcolor{Yellow}(0.312, 0.581) \\
            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.400, 0.630) \\
            & \cellcolor{Yellow}0.60 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.462, 0.687) \\
            & \cellcolor{Yellow}0.80 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.502, 0.727) \\
            & \cellcolor{Red}1.00 & \cellcolor{Red}Bad &
              \cellcolor{Red}(0.318, 0.549) \\
        \hline
        \multirow{5}{*}{0.80}
            & \cellcolor{Green}0.20 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.578, 0.838) \\
            & \cellcolor{Green}0.40 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.736, 0.915) \\
            & \cellcolor{Yellow}0.60 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.587, 0.829) \\
            & \cellcolor{Yellow}0.80 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.599, 0.811) \\
            & \cellcolor{Red}1.00 & \cellcolor{Red}Bad &
              \cellcolor{Red}(0.612, 0.830) \\
        \hline
        \multirow{5}{*}{1.00}
            & \cellcolor{Yellow}0.20 & \cellcolor{Yellow}Good &
              \cellcolor{Yellow}(0.895, 0.999) \\
            & \cellcolor{Yellow}0.40 & \cellcolor{Yellow}Okay &
              \cellcolor{Yellow}(0.927, 0.999) \\
            & \cellcolor{Green}0.60 & \cellcolor{Green}Good &
              \cellcolor{Green}(0.949, 1.000) \\
            & \cellcolor{Yellow}0.80 & \cellcolor{Yellow}Good &
              \cellcolor{Yellow}(0.802, 0.954) \\
            & \cellcolor{Red}1.00 & \cellcolor{Red}Bad &
              \cellcolor{Red}(0.800, 0.955) \\
        \hline
    \end{tabular}
    \end{center}
    \end{table}

    Generally, the model is most effective for lower values of \(\psi\) and
    \(\theta\) but will either converge well or provide useful estimates for the
    vast majority of combinations of \(\psi\) and \(\theta\) values. Twenty-one
    of the
    model's 25 confidence intervals for \(\psi\) include its actual value.
    The model's confidence intervals for each of the
    three values of \(p\) include its actual value for all \(\psi\) and
    \(\theta\) values. On the other hand, the model accurately estimates \(E\)
    for low values of \(\theta\), but consistently overestimates \(E\) when
    \(\psi\geq0.4\) and \(\theta\geq0.6\) regardless of \(E\)'s actual value.
    This may help to explain why the model does not estimate \(\psi\) as
    successfully for higher values of \(\psi\) and \(\theta\).

\section{Spatial correlation of the data}
One approach to dealing with the correlation in the probability of
observing a track in neighboring plots is to use a conditional autoregressive
(CAR) prior distribution for $\theta$, the probability that an otter lays down a
track in a plot on a given day.
